---
title: "T_cell_analysis"
author: "Hongke Peng"
date: "2023-10-25"
output: html_document
---

#load libraries
```{R}
library(Seurat)
library(ggplot2)
library(dittoSeq)
library(harmony)
library(clusterProfiler)
set.seed(1)
```

```{r}
y <- readRDS("make_umap/t_NK.rds")
```

# cell annotation
```{r,fig.width=17,fig.height=5}
c1 <- c(ggsci::pal_npg(alpha = 0.7)(8), ggsci::pal_nejm(alpha = 0.7)(8), ggsci::pal_jama(alpha = 0.7)(8))
y$stage <- "VEN-relapsed"
y$stage[y$patient %in% c("CLL400", "CLL427")] <- "VEN-naive"
p1=DimPlot(y,reduction="wnnUMAP",group.by="stage") + NoAxes()
p2=DimPlot(y,reduction="wnnUMAP",group.by="compartment",pt.size = 0.8, raster = T) + NoAxes()
p3=DimPlot(y,reduction="wnnUMAP",group.by="refined.cluster", label = T, cols = c1, pt.size = 0.8, label.size = 5, order = F, raster = T, label.box = T) + NoAxes()
p <- p3 + p2 + p1
p
```

```{r, fig.width=5.5, fig.height=5}
y$compartment <- factor(y$compartment, levels = c("PB", "LN"))
DimPlot(y,reduction="wnnUMAP",group.by="compartment",pt.size = 0.8, raster = T, cols = c("#3F5386", "#BC3E28")) + NoAxes()
ggsave("Tcell_analysis_figure/umap_by_compartment.png", width = 5.5, height = 5)
```

```{r, fig.width=11, fig.height=5}
y$compartment <- factor(y$compartment, levels = c("PB", "LN"))
DimPlot(y,reduction="wnnUMAP",group.by="compartment",pt.size = 0.8, raster = T, cols = c("#3F5386", "#BC3E28"), split.by = "compartment") + NoAxes()
ggsave("Tcell_analysis_figure/umap_split_by_compartment.png", width = 11, height = 5)
```


```{r, fig.width=8, fig.height=3.5}
df <- table(y$refined.cluster, y$compartment) / rowSums(table(y$refined.cluster, y$compartment))
df <- as.data.frame(df)
colnames(df) <- c("cluster", "compartment", "fraction")
ggplot(df, aes(x=cluster, y=fraction, fill=compartment)) + 
  geom_col() +
  theme_bw() +
  xlab("T cell and NK cell cluster") +
  ylab("Traction of cells from LN and PB") + 
  ggtitle("Cluster composition") + 
  theme(axis.text = element_text(size=12))
ggsave("Tcell_analysis_figure/cluster_composition.pdf", width = 8, height = 3.5)
```



```{r, fig.width=15, fig.height=4}
DimPlot(y,reduction="wnnUMAP",group.by="refined.cluster", label = T, cols = c1, pt.size = 0.8, label.size = 5, split.by = "patient", raster = T) + NoAxes() + NoLegend()
```

#composition analysis
##LN CD4 vs CD8
```{r, fig.width=5, fig.height=2.5}
x <- subset(y, compartment == "LN" & refined.cluster %in% 0:12)
x$cd4_cd8 <- "CD4"
x$cd4_cd8[x$refined.cluster %in% 8:12] <- "CD8"
DimPlot(x,reduction="wnnUMAP",group.by="cd4_cd8", label = T, cols = c1, pt.size = 0.8, label.size = 3, order = F, raster = T) + NoAxes()
df1 <- table(x$patient, x$cd4_cd8) / rowSums(table(x$patient, x$cd4_cd8))
df1 <- as.data.frame(df1)
colnames(df1) <- c("patient", "lineage", "Freq")
df1$stage <- "VEN-naive"
df1$stage[df1$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
df1$stage <- factor(df1$stage, levels = c("VEN-relpase", "VEN-naive"))
df1$patient <- factor(df1$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
df1$lineage <- paste("LN ", df1$lineage, " T")

ggplot(df1, aes(x = patient, y = Freq, fill = stage)) + 
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 8, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_LN_cd4_vs_cd8.pdf", width = 5, height = 2.5)
```

##PB CD4 vs CD8
```{r, fig.width=5, fig.height=2.5}
x <- subset(y, compartment == "PB" & refined.cluster %in% 0:12)
x$cd4_cd8 <- "CD4"
x$cd4_cd8[x$refined.cluster %in% 8:12] <- "CD8"
DimPlot(x,reduction="wnnUMAP",group.by="cd4_cd8", label = T, cols = c1, pt.size = 0.8, label.size = 3, order = F, raster = T) + NoAxes()
df1 <- table(x$patient, x$cd4_cd8) / rowSums(table(x$patient, x$cd4_cd8))
df1 <- as.data.frame(df1)
colnames(df1) <- c("patient", "lineage", "Freq")
df1$stage <- "VEN-naive"
df1$stage[df1$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
df1$stage <- factor(df1$stage, levels = c("VEN-relpase", "VEN-naive"))
df1$patient <- factor(df1$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
df1$lineage <- paste("PB ", df1$lineage, " T")

ggplot(df1, aes(x = patient, y = Freq, fill = stage)) + 
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 8, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_PB_cd4_vs_cd8.pdf", width = 5, height = 2.5)
```
##CD4 LN T cell composition
```{r, fig.width=7.5, fig.height=7.5}
cd4 <- subset(y, refined.cluster %in% 0:7 & compartment == "LN")
LN_vs_PB <- table(cd4$patient, cd4$cell.type)/rowSums(table(cd4$patient, cd4$cell.type))
LN_vs_PB <- as.data.frame(LN_vs_PB)
colnames(LN_vs_PB) <- c("patient", "lineage", "Freq")

LN_vs_PB$stage <- "VEN-naive"
LN_vs_PB$stage[LN_vs_PB$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
LN_vs_PB$stage <- factor(LN_vs_PB$stage, levels = c("VEN-relpase", "VEN-naive"))
LN_vs_PB$patient <- factor(LN_vs_PB$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
#LN_vs_PB$lineage <- paste("LN ", LN_vs_PB$lineage, " T")

ggplot(LN_vs_PB, aes(x = patient, y = Freq, fill = stage)) + 
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  #ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 8, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_LN_cd4_T.pdf", width = 7.5, height = 7.5)
```

##CD8 LN T cell composition
```{r, fig.width=7.5, fig.height=5}
cd8 <- subset(y, refined.cluster %in% 8:12 & compartment == "LN")
LN_vs_PB <- table(cd8$patient, cd8$cell.type)/rowSums(table(cd8$patient, cd8$cell.type))
LN_vs_PB <- as.data.frame(LN_vs_PB)
colnames(LN_vs_PB) <- c("patient", "lineage", "Freq")

LN_vs_PB$stage <- "VEN-naive"
LN_vs_PB$stage[LN_vs_PB$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
LN_vs_PB$stage <- factor(LN_vs_PB$stage, levels = c("VEN-relpase", "VEN-naive"))
LN_vs_PB$patient <- factor(LN_vs_PB$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
#LN_vs_PB$lineage <- paste("LN ", LN_vs_PB$lineage, " T")

ggplot(LN_vs_PB, aes(x = patient, y = Freq, fill = stage)) + 
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  #ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 8, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_LN_cd8_T.pdf", width = 7.5, height = 5)
```

##CD4 PB T cell composition
```{r, fig.width=7.5, fig.height=7.5}
cd4 <- subset(y, refined.cluster %in% 0:7 & compartment == "PB")
LN_vs_PB <- table(cd4$patient, cd4$cell.type)/rowSums(table(cd4$patient, cd4$cell.type))
LN_vs_PB <- as.data.frame(LN_vs_PB)
colnames(LN_vs_PB) <- c("patient", "lineage", "Freq")

LN_vs_PB$stage <- "VEN-naive"
LN_vs_PB$stage[LN_vs_PB$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
LN_vs_PB$stage <- factor(LN_vs_PB$stage, levels = c("VEN-relpase", "VEN-naive"))
LN_vs_PB$patient <- factor(LN_vs_PB$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
LN_vs_PB$lineage <- paste("PB ", LN_vs_PB$lineage, " T")

ggplot(LN_vs_PB, aes(x = patient, y = Freq, fill = stage)) + 
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  #ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 8, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_PB_cd4_T.pdf", width = 7.5, height = 7.5)
```

##CD8 PB T cell composition
```{r, fig.width=7.5, fig.height=5}
cd8 <- subset(y, refined.cluster %in% 8:12 & compartment == "PB")
LN_vs_PB <- table(cd8$patient, cd8$cell.type)/rowSums(table(cd8$patient, cd8$cell.type))
LN_vs_PB <- as.data.frame(LN_vs_PB)
colnames(LN_vs_PB) <- c("patient", "lineage", "Freq")

LN_vs_PB$stage <- "VEN-naive"
LN_vs_PB$stage[LN_vs_PB$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
LN_vs_PB$stage <- factor(LN_vs_PB$stage, levels = c("VEN-relpase", "VEN-naive"))
LN_vs_PB$patient <- factor(LN_vs_PB$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
LN_vs_PB$lineage <- paste("PB ", LN_vs_PB$lineage, " T")

ggplot(LN_vs_PB, aes(x = patient, y = Freq, fill = stage)) + 
  geom_col(position = "dodge", width = 0.5) +
  facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  #ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 8, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_PB_cd8_T.pdf", width = 7.5, height = 5)
```

#Exhausted T cells
##DE between c5, c6
```{r}
DefaultAssay(y) <- "RNA"
x <- subset(y, refined.cluster %in% 5:6)
c6_vs_c5 <- FindMarkers(x, group.by = "stage", ident.1 = "VEN-relapsed", ident.2 = "VEN-naive", test.use = "MAST", logfc.threshold = 0.1)
c6_vs_c5$gene <- rownames(c6_vs_c5)
```

```{r,fig.width=11,fig.height=5}
pt=subset(y, refined.cluster %in% 5:6)
#make sce
matr=pt@assays$RNA@counts
md=pt@meta.data[,c("patient","compartment","stage","refined.cluster")]
sce=SingleCellExperiment(assay=list(counts=matr),colData=md)

pool=scater::aggregateAcrossCells(sce,id=colData(sce)[,c("refined.cluster","patient")])

y=pool
y=DGEList(counts=counts(y),samples=colData(y),remove.zeros=T)
#filtering 
keep=filterByExpr(y,group=y$samples$refined.cluster)
summary(keep)#rm:6329,retain:12523
y=y[keep,,keep.lib.sizes=FALSE]
#calculate normalization factors
y=calcNormFactors(y)
#plot MDS
mds=plotMDS.DGEList(y,plot=F)
gg_df=data.frame(mds[c("x","y")],
                 stage_id=as.character(y$samples$stage),
                 patient_id=y$samples$patient,
                 group_id=y$samples$refined.cluster)

#getPalette=colorRampPalette(brewer.pal(10,"Set1"))
p1=ggplot(gg_df,aes(x,y,col=stage_id,shape=group_id))+
    geom_point(size=4,alpha=0.8)+
    labs(x="MDS dim1",y="MDS dim2")+
    theme(panel.grid.minor=element_blank())+
    #coord_fixed()+
    #ggtitle(paste0(i," MDS plot"))+
    ggtitle("MDS plot: basic clusters") +
    scale_color_manual(values=c1)+
    theme_bw()
p2=ggplot(gg_df,aes(x,y,col=group_id,shape=patient_id))+
    geom_point(size=4,alpha=0.8)+
    labs(x="MDS dim1",y="MDS dim2")+
    theme(panel.grid.minor=element_blank())+
    #coord_fixed()+
    #ggtitle(paste0(i," MDS plot"))+
    ggtitle("MDS plot: basic clusters") +
    scale_color_manual(values=c1)+
    theme_bw()
p <- p1+p2
p
ggsave("Tcell_analysis_figure/MDS_T_C5_C6.pdf", width = 11, height = 5)
```

Q1: C6 vs C5
```{r}
y$samples$refined.cluster <- paste0("C", y$samples$refined.cluster)
#make design
f1=factor(y$samples$refined.cluster,levels=c("C6","C5"))
f2=y$samples$patient
design=model.matrix(~0+f1+f2)
colnames(design)=gsub("f\\d","",colnames(design))
  
#make contrast
contr=makeContrasts(
  comparison=C6-C5,
  levels=colnames(design)
)
  
y=estimateDisp(y,design)
plotBCV(y)
  
fit=glmFit(y,design)
lrt=glmLRT(fit,contrast=contr)
#fit=glmQLFit(y,design)
#lrt=glmQLFTest(fit,contrast=contr)
DEG=topTags(lrt,n=Inf)
  
#the proportion of cell express this gene
df=DEG$table
df$gene <- rownames(df)
write.csv(df, file = "Tcell_analysis_figure/edgeR_C6_vs_C5.csv")
```

```{r,fig.width=7, fig.height=8}
df <- read.csv("Tcell_analysis_figure/edgeR_C6_vs_C5.csv")
EnhancedVolcano::EnhancedVolcano(df,
                x="logFC",y="FDR",
                lab=df$gene,
                #xlim=c(-10,10),
                #ylim=c(0,50),
                title = "CD4 T-cell C6 vs C5",
                subtitle = paste0('FDR cutoff = 0.05', "  logFC cutoff = 1"),
                selectLab = c("LAG3","HAVCR2","BTLA","TCF7", "LEF1", "CCR7", "SELL", "EEF1A1",
                              "GZMA","GZMB","GZMK","GZMH","FASLG","PRF1","GNLY","NKG7","TIMD4"),
                labSize = 6, 
                boxedLabels = T,
                legendPosition = "top",
                legendLabSize = 12,
                legendIconSize = 4.0,
                drawConnectors = T,
                pCutoff = 0.05,
                raster = T,
                FCcutoff = 1)
ggsave("Tcell_analysis_figure/volcano_C5_vs_C6.pdf", width = 7, height = 8)
```

```{r}
up=df[df$FDR<0.05&df$logFC>1,]
go.up=enrichGO(gene=up$gene,
               OrgDb="org.Hs.eg.db",
               keyType="SYMBOL",
               ont="ALL",
               pAdjustMethod="BH",
               pvalueCutoff=0.05,
               qvalueCutoff=0.01)

down=df[df$FDR<0.05&df$logFC<(-1),]
go.down=enrichGO(gene=down$gene,
               OrgDb="org.Hs.eg.db",
               keyType="SYMBOL",
               ont="ALL",
               pAdjustMethod="BH",
               pvalueCutoff=0.05,
               qvalueCutoff=0.01)
```


```{r}
all_gene_set <- data.frame()
library(msigdbr)
##C2 canonical pathway
for (i in c("CP:BIOCARTA", "CP:KEGG", "CP:REACTOME", "CP:WIKIPATHWAYS", "CP:PID")) {
  mtx <- msigdbr(species = "Homo sapiens",category = "C2",subcategory = i)
  mtx <- as.data.frame(mtx)
  all_gene_set <- rbind(all_gene_set, mtx)
}
##Hallmark
mtx <- msigdbr(species = "Homo sapiens",category = "H")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

mtx <- msigdbr(species = "Homo sapiens",category = "C5", subcategory = "GO:BP")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

gmt <- data.frame(term = all_gene_set$gs_name, 
                  gene = all_gene_set$entrez_gene)
#rm genes with small faction of cells expressing it
id=bitr(df$gene,"SYMBOL","ENTREZID","org.Hs.eg.db") #5.38% of input gene IDs are fail to map...
#make geneList
marker=merge(df,id,by.x="gene",by.y="SYMBOL")
marker=data.frame(logFC=marker$logFC,SYMBOL=marker$ENTREZID)
#marker=data.frame(logFC=marker$avg_log2FC,SYMBOL=marker$ENTREZID)
geneList=marker$logFC
names(geneList)=marker$SYMBOL
geneList=sort(geneList,decreasing=T)
gsea.result <- GSEA(geneList, TERM2GENE = gmt, seed = T, pvalueCutoff = 0.01, pAdjustMethod = "BH")
#write.csv(gsea.result@result, "CLL_analysis_figure/gsea_LN_CLL_relpase_vs_pre.csv")
```

##DE between c5, c6
```{r}
DefaultAssay(y) <- "RNA"
x <- subset(y, refined.cluster %in% 8:9)
c6_vs_c5 <- FindMarkers(x, group.by = "stage", ident.1 = "VEN-relapsed", ident.2 = "VEN-naive", test.use = "MAST", logfc.threshold = 0.1)
c6_vs_c5$gene <- rownames(c6_vs_c5)
```

```{r,fig.width=11,fig.height=5}
pt=subset(y, refined.cluster %in% 8:9)
#make sce
matr=pt@assays$RNA@counts
md=pt@meta.data[,c("patient","compartment","stage","refined.cluster")]
sce=SingleCellExperiment(assay=list(counts=matr),colData=md)

pool=scater::aggregateAcrossCells(sce,id=colData(sce)[,c("refined.cluster","patient")])

y=pool
y=DGEList(counts=counts(y),samples=colData(y),remove.zeros=T)
#filtering 
keep=filterByExpr(y,group=y$samples$refined.cluster)
summary(keep)#rm:6329,retain:12523
y=y[keep,,keep.lib.sizes=FALSE]
#calculate normalization factors
y=calcNormFactors(y)
#plot MDS
mds=plotMDS.DGEList(y,plot=F)
gg_df=data.frame(mds[c("x","y")],
                 stage_id=as.character(y$samples$stage),
                 patient_id=y$samples$patient,
                 group_id=y$samples$refined.cluster)

#getPalette=colorRampPalette(brewer.pal(10,"Set1"))
p1=ggplot(gg_df,aes(x,y,col=stage_id,shape=group_id))+
    geom_point(size=4,alpha=0.8)+
    labs(x="MDS dim1",y="MDS dim2")+
    theme(panel.grid.minor=element_blank())+
    #coord_fixed()+
    #ggtitle(paste0(i," MDS plot"))+
    ggtitle("MDS plot: basic clusters") +
    scale_color_manual(values=c1)+
    theme_bw()
p2=ggplot(gg_df,aes(x,y,col=group_id,shape=patient_id))+
    geom_point(size=4,alpha=0.8)+
    labs(x="MDS dim1",y="MDS dim2")+
    theme(panel.grid.minor=element_blank())+
    #coord_fixed()+
    #ggtitle(paste0(i," MDS plot"))+
    ggtitle("MDS plot: basic clusters") +
    scale_color_manual(values=c1)+
    theme_bw()
p <- p1+p2
p
ggsave("Tcell_analysis_figure/MDS_T_C8_C9.pdf", width = 11, height = 5)
```

Q1: C6 vs C5
```{r}
y$samples$refined.cluster <- paste0("C", y$samples$refined.cluster)
#make design
f1=factor(y$samples$refined.cluster,levels=c("C8","C9"))
f2=y$samples$patient
design=model.matrix(~0+f1+f2)
colnames(design)=gsub("f\\d","",colnames(design))
  
#make contrast
contr=makeContrasts(
  comparison=C8-C9,
  levels=colnames(design)
)
  
y=estimateDisp(y,design)
plotBCV(y)
  
fit=glmFit(y,design)
lrt=glmLRT(fit,contrast=contr)
#fit=glmQLFit(y,design)
#lrt=glmQLFTest(fit,contrast=contr)
DEG=topTags(lrt,n=Inf)
  
#the proportion of cell express this gene
df=DEG$table
df$gene <- rownames(df)
write.csv(df, file = "Tcell_analysis_figure/edgeR_C8_vs_C9.csv")
```

```{r,fig.width=7, fig.height=8}
df <- read.csv("Tcell_analysis_figure/edgeR_C8_vs_C9.csv")
EnhancedVolcano::EnhancedVolcano(df,
                x="logFC",y="FDR",
                lab=df$gene,
                #xlim=c(-10,10),
                #ylim=c(0,50),
                title = "CD8 T-cell C8 vs C9",
                subtitle = paste0('FDR cutoff = 0.05', "  logFC cutoff = 1"),
                selectLab = c("LAG3","HAVCR2","BTLA","TCF7", "CCR7", "SELL", "EEF1A1",
                              "GZMA","GZMB","GZMH","FASLG","PRF1","GNLY","NKG7","TIMD4"),
                labSize = 6, 
                boxedLabels = T,
                legendPosition = "top",
                legendLabSize = 12,
                legendIconSize = 4.0,
                drawConnectors = T,
                pCutoff = 0.05,
                raster = T,
                FCcutoff = 1)
ggsave("Tcell_analysis_figure/volcano_C8_vs_C9.pdf", width = 7, height = 8)
```

```{r}
up=df[df$FDR<0.05&df$logFC>1,]
go.up=enrichGO(gene=up$gene,
               OrgDb="org.Hs.eg.db",
               keyType="SYMBOL",
               ont="ALL",
               pAdjustMethod="BH",
               pvalueCutoff=0.05,
               qvalueCutoff=0.01)

down=df[df$FDR<0.05&df$logFC<(-1),]
go.down=enrichGO(gene=down$gene,
               OrgDb="org.Hs.eg.db",
               keyType="SYMBOL",
               ont="ALL",
               pAdjustMethod="BH",
               pvalueCutoff=0.05,
               qvalueCutoff=0.01)
```


```{r}
all_gene_set <- data.frame()
library(msigdbr)
##C2 canonical pathway
for (i in c("CP:BIOCARTA", "CP:KEGG", "CP:REACTOME", "CP:WIKIPATHWAYS", "CP:PID")) {
  mtx <- msigdbr(species = "Homo sapiens",category = "C2",subcategory = i)
  mtx <- as.data.frame(mtx)
  all_gene_set <- rbind(all_gene_set, mtx)
}
##Hallmark
mtx <- msigdbr(species = "Homo sapiens",category = "H")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

mtx <- msigdbr(species = "Homo sapiens",category = "C5", subcategory = "GO:BP")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

gmt <- data.frame(term = all_gene_set$gs_name, 
                  gene = all_gene_set$entrez_gene)
#rm genes with small faction of cells expressing it
id=bitr(df$gene,"SYMBOL","ENTREZID","org.Hs.eg.db") #5.38% of input gene IDs are fail to map...
#make geneList
marker=merge(df,id,by.x="gene",by.y="SYMBOL")
marker=data.frame(logFC=marker$logFC,SYMBOL=marker$ENTREZID)
#marker=data.frame(logFC=marker$avg_log2FC,SYMBOL=marker$ENTREZID)
geneList=marker$logFC
names(geneList)=marker$SYMBOL
geneList=sort(geneList,decreasing=T)
gsea.result <- GSEA(geneList, TERM2GENE = gmt, seed = T, pvalueCutoff = 0.01, pAdjustMethod = "BH")
#write.csv(gsea.result@result, "CLL_analysis_figure/gsea_LN_CLL_relpase_vs_pre.csv")
```

##C6 VEN-relapsed vs VEN-naive

```{r}
pt=subset(y, refined.cluster == 5)
#make sce
matr=pt@assays$RNA@counts
md=pt@meta.data[,c("patient","stage")]
sce=SingleCellExperiment(assay=list(counts=matr),colData=md)

pool=scater::aggregateAcrossCells(sce,id=colData(sce)[,c("stage","patient")])

y=pool
y=DGEList(counts=counts(y),samples=colData(y),remove.zeros=T)
#filtering 
keep=filterByExpr(y,group=y$samples$stage)
summary(keep)#rm:6329,retain:12523
y=y[keep,,keep.lib.sizes=FALSE]
#calculate normalization factors
y=calcNormFactors(y)
#plot MDS
mds=plotMDS.DGEList(y,plot=F)
gg_df=data.frame(mds[c("x","y")],
                 stage_id=as.character(y$samples$stage),
                 patient_id=y$samples$patient)

#getPalette=colorRampPalette(brewer.pal(10,"Set1"))
ggplot(gg_df,aes(x,y,col=stage_id))+
    geom_point(size=4,alpha=0.8)+
    labs(x="MDS dim1",y="MDS dim2")+
    theme(panel.grid.minor=element_blank())+
    #coord_fixed()+
    #ggtitle(paste0(i," MDS plot"))+
    ggtitle("MDS plot: basic clusters") +
    scale_color_manual(values=c1)+
    theme_bw()
```
```{r}
y$samples$stage <- gsub("-", "_", y$samples$stage)
#make design
f1=factor(y$samples$stage,levels=c("VEN_relapsed","VEN_naive"))
design=model.matrix(~0+f1)
colnames(design)=gsub("f\\d","",colnames(design))
  
#make contrast
contr=makeContrasts(
  comparison=VEN_relapsed-VEN_naive,
  levels=colnames(design)
)
  
y=estimateDisp(y,design)
plotBCV(y)
  
fit=glmFit(y,design)
lrt=glmLRT(fit,contrast=contr)
#fit=glmQLFit(y,design)
#lrt=glmQLFTest(fit,contrast=contr)
DEG=topTags(lrt,n=Inf)
  
#the proportion of cell express this gene
df=DEG$table
df$gene <- rownames(df)
write.csv(df, file = "Tcell_analysis_figure/edgeR_C6_vs_C5.csv")
```

```{r,fig.width=7, fig.height=8}
EnhancedVolcano::EnhancedVolcano(df,
                x="logFC",y="FDR",
                lab=rownames(df),
                #xlim=c(-10,10),
                #ylim=c(0,50),
                title = "CD4 T-cell C6 vs C5",
                subtitle = paste0('FDR cutoff = 0.05', "  logFC cutoff = 1"),
                #selectLab = c("LAG3","HAVCR2","BTLA","TCF7", "LEF1", "CCR7", "GZMK", "SELL", "EEF1A1"),
                labSize = 4,
                legendPosition = "top",
                legendLabSize = 12,
                legendIconSize = 4.0,
                drawConnectors = T,
                pCutoff = 0.05,
                raster = T,
                FCcutoff = 1)
ggsave("Tcell_analysis_figure/volcano_C5_vs_C6.pdf", width = 7, height = 8)
```


It seems like that the DEG found here were also observed in CLL comparison
```{r}
CLL_compare <- read.csv("CLL_analysis_figure/edgeR_LN_CLL_relapse_vs_pre.csv", row.names = 1)
CLL_compare <- CLL_compare$gene[CLL_compare$FDR < 0.05]
T_compare <- df$gene[df$FDR < 0.05]
intersect(T_compare, CLL_compare)
df <- df[!(df$gene %in% intersect(T_compare, CLL_compare)), ]
counts.f <- y$counts[!(rownames(y$counts) %in% intersect(T_compare, CLL_compare)), ]
y$counts <- counts.f
```


```{r}
all_gene_set <- data.frame()
library(msigdbr)
##C2 canonical pathway
for (i in c("CP:BIOCARTA", "CP:KEGG", "CP:REACTOME", "CP:WIKIPATHWAYS", "CP:PID")) {
  mtx <- msigdbr(species = "Homo sapiens",category = "C2",subcategory = i)
  mtx <- as.data.frame(mtx)
  all_gene_set <- rbind(all_gene_set, mtx)
}
##Hallmark
mtx <- msigdbr(species = "Homo sapiens",category = "H")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

mtx <- msigdbr(species = "Homo sapiens",category = "C5", subcategory = "GO:BP")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

gmt <- data.frame(term = all_gene_set$gs_name, 
                  gene = all_gene_set$entrez_gene)
#rm genes with small faction of cells expressing it
id=bitr(df$gene,"SYMBOL","ENTREZID","org.Hs.eg.db") #5.38% of input gene IDs are fail to map...
#make geneList
marker=merge(df,id,by.x="gene",by.y="SYMBOL")
marker=data.frame(logFC=marker$logFC,SYMBOL=marker$ENTREZID)
#marker=data.frame(logFC=marker$avg_log2FC,SYMBOL=marker$ENTREZID)
geneList=marker$logFC
names(geneList)=marker$SYMBOL
geneList=sort(geneList,decreasing=T)
gsea.result <- GSEA(geneList, TERM2GENE = gmt, seed = T, pvalueCutoff = 0.01, pAdjustMethod = "BH")
#write.csv(gsea.result@result, "CLL_analysis_figure/gsea_LN_CLL_relpase_vs_pre.csv")
```



##publish Tex markers
```{r, fig.width=7, fig.height=9}
y <- subset(y, refined.cluster %in% c(5,6,8,9))
#read in signatures
DefaultAssay(y) <- "RNA"
cd4_tex <- read.csv("Tcell_analysis_figure/CD4_Tex_marker.csv")
cd4_tex <- list(cd4_tex$Gene.Symbol)
y <- AddModuleScore(y, features = cd4_tex, ctrl = 5, name = "CD4_Tex_score")

cd8_tex <- read.csv("Tcell_analysis_figure/CD8_Tex_marker.csv")
cd8_tex <- list(cd8_tex$Gene.Symbol)
y <- AddModuleScore(y, features = cd8_tex, ctrl = 5, name = "CD8_Tex_score")

#Tex_score <- list(unique(c(cd4_tex$Gene.Symbol, cd8_tex$Gene.Symbol)))
#y <- AddModuleScore(y, features = Tex_score, ctrl = 5, name = "Tex_score")

cd8_pre_ex <- read.csv("Tcell_analysis_figure/CD8_T_markers.csv")
pre_marker1 <- cd8_pre_ex[cd8_pre_ex$Cluster == "CD8-C4-GZMK", ]
pre_marker1 <- list(pre_marker1$Gene.Symbol)
y <- AddModuleScore(y, features = pre_marker1, ctrl = 5, name = "pre_marker")

cd8_pre_ex <- read.csv("Tcell_analysis_figure/CD8_T_markers.csv")
pre_marker2 <- cd8_pre_ex[cd8_pre_ex$Cluster == "CD8-C5-ZNF683", ]
pre_marker2 <- list(pre_marker2$Gene.Symbol)
y <- AddModuleScore(y, features = pre_marker1, ctrl = 100, name = "pre_marker_2")

cytotoxity <- list(c("PRF1", "IFNG", "GNLY", "NKG7", "GZMB", "GZMA", "GZMH", "KLRK1", "KLRD1", "CTSW", "CST7"))
y <- AddModuleScore(y, features = cytotoxity, ctrl = 10, name = "cytotoxity")

naiveness <- list(c("CCR7", "TCF7", "LEF1", "SELL"))
y <- AddModuleScore(y, features = naiveness, ctrl = 5, name = "naiveness")

#stem_Tex <- list(c("PDCD1","CTLA4","TOX","TOX2","TIGIT","CCR7","LEF1","TCF7","SELL","EEF1A1","IL6R","IGFBP4","IGFL2"))
#y <- AddModuleScore(y, features = stem_Tex, ctrl = 5, name = "stem_like_Tex")

Treg_signature <- list(c("FOXP3","IL2RA","ICOS","IKZF2","CTLA4"))
y <- AddModuleScore(y, features = Treg_signature, ctrl = 5, name = "Treg_signature")
another_marker <- read.csv("Tcell_analysis_figure/another_markers.csv")
pre_marker1 <- list(another_marker$CD8_Tpex)
y <- AddModuleScore(y, features = pre_marker1, ctrl = 100, name = "pre_marker")

p1 <- VlnPlot(y, features = "CD8_Tex_score1", group.by = "refined.cluster", cols = c1, pt.size = 0) + NoLegend()
p2 <- VlnPlot(y, features = "pre_marker1", group.by = "refined.cluster", cols = c1, pt.size = 0) + NoLegend()
p3 <- VlnPlot(y, features = "cytotoxity1", group.by = "refined.cluster", cols = c1, pt.size = 0) + NoLegend()
p4 <- VlnPlot(y, features = "Treg_signature1", group.by = "refined.cluster", cols = c1, pt.size = 0) + NoLegend()
p5 <- VlnPlot(y, features = "CD4_Tex_score1", group.by = "refined.cluster", cols = c1, pt.size = 0) + NoLegend()
p <- p1 + p5 + p2 + p3 + p4
p
ggsave(plot = p, filename = "Tcell_analysis_figure/vlnplot_scores.pdf", width = 6, height = 9)
```

```{r, fig.width=20, fig.height=5}
VlnPlot(y, features = "pre_marker1", group.by = "refined.cluster", cols = c1, pt.size = 0, split.by = "stage")
```

```{r, fig.width=20, fig.height=5}
VlnPlot(y, features = "CD8_Tex_score1", group.by = "refined.cluster", cols = c1, pt.size = 0, split.by = "stage")
```

```{r, fig.width=20, fig.height=5}
VlnPlot(y, features = "CD4_Tex_score1", group.by = "refined.cluster", cols = c1, pt.size = 0, split.by = "stage")
```

```{r, fig.width=5, fig.height=9}
DefaultAssay(y) <- "DSB"
rownames(y)[grep("CXCR5", rownames(y))]
VlnPlot(y, features = c("rna_GZMB", "rna_GZMA","rna_GZMK", "rna_PRF1","rna_GNLY","rna_IFNG","rna_TOX","rna_PDCD1","rna_CTLA4","rna_TIGIT", "rna_LAG3"), group.by = "refined.cluster", cols = c1, pt.size = 0) + NoLegend()
VlnPlot(y, features = c("rna_GZMK", "rna_GZMB", "rna_GZMA", "rna_GZMH", "rna_PRF1"), group.by = "refined.cluster", cols = c1, pt.size = 0, ncol = 1, split.by = "stage") + NoLegend()
ggsave("Tcell_analysis_figure/vlnplot_GZMK.pdf", width = 5, height = 9)
```
```{r,fig.width=15, fig.height=10}
DefaultAssay(y) <- "RNA"
FeaturePlot(y, reduction = "wnnUMAP", features = c("GZMK", "GZMB", "GZMA", "GZMH", "PRF1"), min.cutoff = "q1", max.cutoff = "q99", ncol = 3, order = F, raster = T)
ggsave("Tcell_analysis_figure/dimplot_Tcell_tytoxicity.pdf", width = 15, height = 10)
```


```{r, fig.width=5, fig.height=3}
x <- subset(y, refined.cluster %in% c(5, 6, 8, 9))
DefaultAssay(x) <- "RNA"
stem_Tex <- list(c("BTLA","FOXP3","TNFRSF4","CCR7","TCF7","SELL","EEF1A1","IL6R","IGFBP4","IGFL2"))
x <- AddModuleScore(x, features = stem_Tex, ctrl = 5, name = "stem_like_Tex")
VlnPlot(x, features = "stem_like_Tex1", group.by = "refined.cluster", cols = c1[c(6,7,9,10)], pt.size = 0) + NoLegend()
ggsave("Tcell_analysis_figure/vlnplot_stem_like_Tex_score.pdf", width = 5, height = 3)
```
```{r, fig.width=4, fig.height=6}
x <- subset(y, refined.cluster %in% c(5, 6, 8, 9))
DefaultAssay(x) <- "RNA"
marker_gene=c("CD4","CD8A",#lineage
              "CCR7","LEF1","TCF7","SELL","EEF1A1","IL6R","IGFBP4","IGFL2",#naiveness
              "GNLY","IFNG","NKG7","PRF1","GZMA","GZMB","GZMH","GZMK","GZMM","CST7",
              "TNFRSF9","LAG3","HAVCR2","BTLA", "TOX", "TOX2")
x$refined.cluster <- factor(y$refined.cluster, levels = c(5, 6, 8, 9))
DotPlot(x,assay="RNA",cols="RdBu",features=marker_gene,cluster.idents=T, group.by = "refined.cluster") + 
  theme(axis.text.x=element_text(size=15)) + 
  xlab(NULL) +ylab(NULL) + ggtitle("Tex gene signatures") +
  coord_flip()
ggsave("Tcell_analysis_figure/dotplot_Tex_signatures.pdf", width = 4, height = 6)
```
##LN CD4 cytotoxic T fraction
```{r, fig.width=6, fig.height=5}
cd4 <- subset(y, refined.cluster %in% 0:7 & compartment == "LN")
cd4$subtype <- cd4$cell.type
cd4$subtype[cd4$refined.cluster %in% 3] <- "Tfh"
cd4$subtype[cd4$refined.cluster %in% 4] <- "Treg"
cd4$subtype[cd4$refined.cluster %in% 5:6] <- "CD4_GZMK"

LN_vs_PB <- table(cd4$patient, cd4$subtype)/rowSums(table(cd4$patient, cd4$subtype))
LN_vs_PB <- as.data.frame(LN_vs_PB)
colnames(LN_vs_PB) <- c("patient", "lineage", "Freq")

LN_vs_PB$stage <- "VEN-naive"
LN_vs_PB$stage[LN_vs_PB$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
LN_vs_PB$stage <- factor(LN_vs_PB$stage, levels = c("VEN-relpase", "VEN-naive"))
LN_vs_PB$patient <- factor(LN_vs_PB$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
#LN_vs_PB$lineage <- paste("LN ", LN_vs_PB$lineage, " T")

ggplot(LN_vs_PB, aes(x = patient, y = Freq, fill = lineage)) + 
  geom_col(width = 0.8) +
  #facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  #ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 12, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_LN_cd4_GZMK_T.pdf", width = 6, height = 5)
```

```{r, fig.width=6, fig.height=5}
cd4 <- subset(y, refined.cluster %in% 0:7 & compartment == "LN")
cd4$refined.cluster <- as.numeric(cd4$refined.cluster)
LN_vs_PB <- table(cd4$patient, cd4$refined.cluster)/rowSums(table(cd4$patient, cd4$refined.cluster))
LN_vs_PB <- as.data.frame(LN_vs_PB)
colnames(LN_vs_PB) <- c("patient", "lineage", "Freq")

LN_vs_PB$stage <- "VEN-naive"
LN_vs_PB$stage[LN_vs_PB$patient %in% c("CLL281", "CLL295")] <- "VEN-relpase"
LN_vs_PB$stage <- factor(LN_vs_PB$stage, levels = c("VEN-relpase", "VEN-naive"))
LN_vs_PB$patient <- factor(LN_vs_PB$patient, levels = c("CLL400", "CLL427", "CLL281", "CLL295"))
#LN_vs_PB$lineage <- paste("LN ", LN_vs_PB$lineage, " T")

ggplot(LN_vs_PB, aes(x = patient, y = Freq, fill = lineage)) + 
  geom_col(width = 0.8) +
  #facet_wrap(~lineage) +
  theme_bw() +
  xlab(NULL) +
  ylab("Fraction") + 
  #ylim(c(0, 0.8)) + 
  scale_fill_manual(values = c1) + 
  theme(axis.text.x = element_text(size = 12, hjust = 1, angle = 45), 
        axis.text.y = element_text(size = 12))
ggsave("Tcell_analysis_figure/composition_LN_cd4_T_patients.pdf", width = 6, height = 5)
```

```{r}
all_gene_set <- data.frame()
library(msigdbr)
##C2 canonical pathway
for (i in c("CP:BIOCARTA", "CP:KEGG", "CP:REACTOME", "CP:WIKIPATHWAYS", "CP:PID")) {
  mtx <- msigdbr(species = "Homo sapiens",category = "C2",subcategory = i)
  mtx <- as.data.frame(mtx)
  all_gene_set <- rbind(all_gene_set, mtx)
}
##Hallmark
mtx <- msigdbr(species = "Homo sapiens",category = "H")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

mtx <- msigdbr(species = "Homo sapiens",category = "C5", subcategory = "GO:BP")
mtx <- as.data.frame(mtx)
all_gene_set <- rbind(all_gene_set, mtx)

gmt <- data.frame(term = all_gene_set$gs_name, 
                  gene = all_gene_set$gene_symbol)

unique(gmt$term[grep("CYTOKINE", gmt$term)])
cytokine <- list(gmt$gene[gmt$term == "REACTOME_CYTOKINE_SIGNALING_IN_IMMUNE_SYSTEM"])
y <- AddModuleScore(y, features = cytokine, ctrl = 100, name = "cytokine")
VlnPlot(y, features = "cytokine1", group.by = "refined.cluster", cols = c1, pt.size = 0)
```



##CD4 exhausted cells
```{r, fig.width=15, fig.height=5}
x <- subset(y, refined.cluster %in% 5:6)
for( i in Reductions(x)) {
  x[[i]] <- NULL
}

DefaultAssay(x) <- "RNA"
#x <- NormalizeData(x)
#x <- FindVariableFeatures(x, nfeatures = 2000)
#x = ScaleData(x, vars.to.regress=c("subsets_Mito_percent"))
#x = ScaleData(x)
#runPCA
x <- SCTransform(x)
hvg <- VariableFeatures(x)
hvg <- hvg[!grepl("^TRBV", hvg)]
hvg <- hvg[!grepl("^TRAV", hvg)]
hvg <- hvg[!grepl("^TRDV", hvg)]
#hvg <- hvg[!grepl("^TRBV", hvg)]
#hvg <- hvg[!grepl("^TRAV", hvg)]
x = RunPCA(x, reduction.name = "uncorrPCA", reduction.key="uncorrPCA_", features = hvg)
#run harmony
x$stage <- "treatment-naive"
x$stage[x$patient %in% c("CLL400", "CLL427")] <- "VEN-naive"
x$stage[x$patient %in% c("CLL281", "CLL295")] <- "VEN-relapsed"
theta.to.use=2
x=RunHarmony(x,
               group.by.vars=c("patient"),
               reduction="uncorrPCA",
               theta=2,
               plot_convergence=T,
               reduction.save ="harmonyPCA",
               kmeans_init_nstart=20, 
               kmeans_init_iter_max=100, 
               assay.use = "SCT"
             )
ElbowPlot(object=x,ndims=50,reduction="uncorrPCA")
ElbowPlot(object=x,ndims=50,reduction="harmonyPCA")
#x <- RunUMAP(x, reduction = "uncorrPCA", dims = 1:20, reduction.name = "uncorrUMAP")
#x <- FindNeighbors(x, reduction = "uncorrPCA", dims = 1:20, graph.name = "u")
#x <- FindClusters(x, graph.name = "u", resolution = 0.5)
#p1 <- dittoDimPlot(x, reduction.use = "uncorrUMAP", var = "compartment", size = 0.5, do.raster = T, do.label = T, labels.size = 3, order = "increasing")
#p2 <- dittoDimPlot(x, reduction.use = "uncorrUMAP", var = "u_res.0.5", size = 0.5, do.raster = T, do.label = T, labels.size = 3, order = "increasing")
#p3 <- dittoDimPlot(x, reduction.use = "uncorrUMAP", var = "patient", size = 0.5, do.raster = F)
#p1 + p2 + p3

x <- RunUMAP(x, reduction = "harmonyPCA", dims = 1:20, reduction.name = "harmonyUMAP", min.dist = 0.01)
x <- FindNeighbors(x, reduction = "harmonyPCA", dims = 1:20, graph.name = "h")
x <- FindClusters(x, graph.name = "h", resolution = 0.5)
p1=DimPlot(x,reduction="harmonyUMAP",group.by="patient",cols=dittoColors(), pt.size = 2, raster = T) + NoAxes()
p2=DimPlot(x,reduction="harmonyUMAP",group.by="refined.cluster",pt.size = 2, raster = T) + NoAxes()
p3=DimPlot(x,reduction="harmonyUMAP",group.by="h_res.0.5", label = T, label.box = T, pt.size = 2, label.size = 6, order = F, raster = T, cols = c1) + NoAxes()
p <- p3 + p2 + p1
p
```


```{r, fig.width=7, fig.height=3}
#read in signatures
DefaultAssay(x) <- "RNA"
cd4_tex <- read.csv("Tcell_analysis_figure/CD4_Tex_marker.csv")
cd4_tex <- list(cd4_tex$Gene.Symbol)
x <- AddModuleScore(x, features = cd4_tex, ctrl = 100, name = "CD4_Tex_score")

cd8_tex <- read.csv("Tcell_analysis_figure/CD8_Tex_marker.csv")
cd8_tex <- list(cd8_tex$Gene.Symbol)
x <- AddModuleScore(x, features = cd8_tex, ctrl = 100, name = "CD8_Tex_score")

cd8_pre_ex <- read.csv("Tcell_analysis_figure/CD8_T_markers.csv")
pre_marker1 <- cd8_pre_ex[cd8_pre_ex$Cluster == "CD8-C4-GZMK", ]
pre_marker1 <- list(pre_marker1$Gene.Symbol)
x <- AddModuleScore(x, features = pre_marker1, ctrl = 100, name = "pre_marker")

cytotoxity <- list(c("PRF1", "IFNG", "GNLY", "NKG7", "GZMB", "GZMA", "GZMH", "KLRK1", "KLRB1", "KLRD1", "CTSW", "CST7"))
x <- AddModuleScore(x, features = cytotoxity, ctrl = 100, name = "cytotoxity")

naiveness <- list(c("CCR7", "TCF7", "LEF1", "SELL"))
x <- AddModuleScore(x, features = naiveness, ctrl = 100, name = "naiveness")

#pre_marker1 <- list(another_marker$CD8_Tex)
#y <- AddModuleScore(x, features = pre_marker1, ctrl = 100, name = "pre_marker")

#another_marker <- read.csv("Tcell_analysis_figure/another_markers.csv")

VlnPlot(x, features = "CD4_Tex_score1", group.by = "h_res.0.5", cols = c1, pt.size = 0)
VlnPlot(x, features = "CD8_Tex_score1", group.by = "h_res.0.5", cols = c1, pt.size = 0)
VlnPlot(x, features = "pre_marker1", group.by = "h_res.0.5", cols = c1, pt.size = 0)
VlnPlot(x, features = "cytotoxity1", group.by = "h_res.0.5", cols = c1, pt.size = 0)
VlnPlot(x, features = "rna_CD4", group.by = "h_res.0.5", cols = c1, pt.size = 0)
```


```{r, fig.width=12, fig.height=12}
DefaultAssay(x) <- "DSB"
rownames(x)[grep("CXCR5", rownames(x))]
FeaturePlot(x, reduction = "harmonyUMAP", features = c("CD279-(PD-1)","TIGIT-(VSTM3)","CD223-(LAG-3)","CD366-(Tim-3)","CD152-(CTLA-4)", "rna_GZMB", "rna_GZMA", "rna_GZMH", "rna_TOX"), min.cutoff = "q2", max.cutoff = "q99", ncol = 3, pt.size = 0.1, order = T)
```

##no end







































































































#end