---
title: "overall_umap"
author: "Hongke Peng"
date: "16/12/2021"
output: html_document
---

#load packages
```{R}
#library(tidyr)
library(dsb)
library(dplyr)
library(Seurat)
library(ggplot2)
library(harmony)
```

#load data
```{r}
srt=readRDS("for_paper_data/srt_all.rds")
srt[["percent.mt"]]=PercentageFeatureSet(srt,pattern="^MT-")

table(srt$Sample)
srt$stage="post"
srt$stage[grepl("C1D1",srt$Sample)]="pre"
table(srt$stage)
```

#part 1 -----------------
#normalization,scale,remove BCR/TCR genes and batch correction
```{r}
#normalize and scale on RNA
DefaultAssay(srt)="RNA"
srt=NormalizeData(srt)
srt=FindVariableFeatures(srt,nfeatures=3000)
hvg=VariableFeatures(srt)[!grepl("^IGHV",VariableFeatures(srt))]#exclude IGHV genes
hvg=hvg[!grepl("^IGLV",hvg)]#exclude IGLV genes
hvg=hvg[!grepl("^IGKV",hvg)]
hvg=hvg[!grepl("^TRAV",hvg)]
hvg=hvg[!grepl("^TRBV",hvg)]
hvg=hvg[!grepl("^TRGV",hvg)]
srt=ScaleData(srt)
#srt=ScaleData(srt,vars.to.regress=c("S.Score","G2M.Score","percent.mt"))
#runPCA
#srt=RunPCA(srt,features=hvg,reduction.name="uncorrPCA",reduction.key="uncorrPCA_")
srt=RunPCA(srt,reduction.name="uncorrPCA",reduction.key="uncorrPCA_")
ElbowPlot(object=srt,ndims=50,reduction="uncorrPCA")
#run harmony
theta.to.use=1
srt=RunHarmony(srt,
               group.by.vars="patient",
               reduction="uncorrPCA",
               theta=theta.to.use,
               plot_convergence=T,
               reduction.save="harmonyPCA",
               kmeans_init_nstart=20,
               kmeans_init_iter_max=100)
ElbowPlot(object=srt,ndims=50,reduction="harmonyPCA")
```
#run umap
```{r}
srt=RunUMAP(srt,
            dims=1:30,
            reduction="harmonyPCA",
            reduction.name="harmonyUMAP",
            reduction.key="harmonyUMAP_")
srt=FindNeighbors(srt,
                  reduction="harmonyPCA",
                  dims=1:30,
                  graph.name="h")
srt=FindClusters(srt,graph.name="h",algorithm=1,resolution=0.5)
```

```{r,fig.width=15,fig.height=5}
p1=DimPlot(srt,reduction="harmonyUMAP",group.by="patient",order=T) + NoAxes()
p2=DimPlot(srt,reduction="harmonyUMAP",group.by="stage",order=F) + NoAxes()
p3=DimPlot(srt,reduction="harmonyUMAP",order=T,label=T) + NoAxes()
p=p1+p2+p3
ggsave(plot=p,filename="overall_umap_figure/umap.pdf",width=15,height=5)
p
```

#part 2 -----------------
analysis in this part shows that: 
in *h_res.0.5 clusters*: 
- cluster 10: dead cells
- cluster 13: doublets
- cluster 14: proliferating cells
- cluster 9: doublets

#find lowq cell and doublets
```{r,fig.width=10,fig.height=10}
for (i in seq(0,2,0.25)) {
  srt=FindClusters(srt,graph.name="h",algorithm=1,resolution=i)
}
library(clustree)
clustree::clustree(srt,prefix="h_res.")
ggsave("overall_umap_figure/clustree.pdf",width=15,height=15)
```

```{r,fig.width=10,fig.height=5}
DimPlot(srt,reduction="harmonyUMAP",order=T,label=T,split.by="stage",group.by="h_res.0.5") + NoAxes()
ggsave("overall_umap_figure/dimplot_split_by_stage.pdf",width=10,height=5)
```

```{r,fig.width=5,fig.height=5}
DimPlot(srt,reduction="harmonyUMAP",order=T,label=T,group.by="h_res.0.5") + NoAxes()
```

```{r,fig.width=20,fig.height=10}
VlnPlot(srt,features=c("nCount_RNA","nFeature_RNA","percent.mt","S.Score"),pt.size=0,ncol=2,group.by="h_res.0.5")
ggsave("overall_umap_figure/vlnplot_basis.pdf",width=20,heigh)
```
```{r,fig.width=15,fig.height=15}
DefaultAssay(srt)="DSB"
rownames(srt)[grep("CD1",rownames(srt))]
FeaturePlot(srt,reduction="harmonyUMAP",features=c("CD3-UCHT1","CD4-RPA-T4","CD8","CD19.1","CD20","CD1d","CD56-(NCAM)","CD11c","CD16"),min.cutoff="q2",max.cutoff="q98",ncol=3,order=T)
ggsave("overall_umap_figure/featureplot_surface_marker.pdf",width=15,height=15)
```

```{r,fig.width=15,fig.height=10}
DefaultAssay(srt)="RNA"
rownames(srt)[grep("XCL",rownames(srt))]
FeaturePlot(srt,reduction="harmonyUMAP",features=c("S.Score","MKI67","STMN1","TUBA1B"),min.cutoff="q2",max.cutoff="q98",ncol=3,order=T)
ggsave("overall_umap_figure/featureplot_prolifering_cells.pdf",width=15,height=10)
```
analysis in this part shows that: 
in *h_res.0.5 clusters*: 
- cluster 10: dead cells
- cluster 13: doublets
- cluster 14: proliferating cells
- cluster 9: doublets

```{r}
saveRDS(srt,"for_paper_data/srt_filtered.rds")
```

```{r,fig.width=5,fig.height=5}
srt$seurat_clusters=srt$h_res.0.5
to_rm=names(srt@meta.data)[grep("h_res.",names(srt@meta.data))]
for (i in to_rm) {
  srt[[i]]=NULL
}
srt@active.ident=srt$seurat_clusters
srt=subset(srt,seurat_clusters %in% c(0:8,11,12,15))
DimPlot(srt,reduction="harmonyUMAP",label=T) + NoAxes()
ggsave("overall_umap_figure/dimplot_no_lowq_doublets_umap.pdf",width=5,height=5)
```
#make subsets data
```{r}
saveRDS(srt,"for_paper_data/srt_filtered.rds")
```

```{r}
srt=readRDS("for_paper_data/srt_filtered.rds")
t=subset(srt,seurat_clusters %in% c(0,1,2))
t_nk=subset(srt,seurat_clusters %in% c(0,1,2,3,8))
b=subset(srt,seurat_clusters %in% c(4,6,7))
mono=subset(srt,seurat_clusters %in% c(5,11,12,15))

saveRDS(t,"for_paper_data/raw_t.rds")
saveRDS(t_nk,"for_paper_data/raw_t_nk.rds")
saveRDS(b,"for_paper_data/raw_b.rds")
saveRDS(mono,"for_paper_data/raw_mono.rds")
```

#part 3 -----------------
#save data
```{r}
saveRDS(srt,"for_paper_data/srt_all.rds")
```

#make subsets data


```{r}
t_nk=subset(srt,seurat_clusters %in% c(0,1,2,3,8))
b=subset(srt,seurat_clusters %in% c(4,6,7))
mono=subset(srt,seurat_clusters %in% c(5,11,12,15))

saveRDS(t_nk,"for_paper_data/raw_t_nk.rds")
saveRDS(b,"for_paper_data/raw_b.rds")
saveRDS(mono,"for_paper_data/raw_mono.rds")
```


#part 4 -----------------
```{r,fig.width=5,fig.height=5}
srt=readRDS("for_paper_data/srt_filtered.rds")
DimPlot(srt,reduction="harmonyUMAP",label=T) + NoAxes()
```

#1st normalize, scale data
```{r,fig.width=10,fig.height=10}
#remove all the reduction object
for (i in names(srt@reductions)) {
  srt[[i]]=NULL
}
DefaultAssay(srt)="RNA"
srt=NormalizeData(srt)
srt=FindVariableFeatures(srt,nfeatures=2000)
#t=ScaleData(t,vars.to.regress=c("S.Score","G2M.Score","percent.mt"))
srt=ScaleData(srt)
hvg=VariableFeatures(srt)
srt=RunPCA(srt,features=hvg,reduction.name="uncorrPCA",reduction.key="uncorrPCA_")
srt=RunHarmony(srt,theta=1,
             group.by.vars="patient",
             reduction="uncorrPCA",
             reduction.save="harmonyPCA",
             kmeans_init_nstart=20,
             kmeans_init_iter_max=100,
             plot_convergence=T)
ElbowPlot(srt,reduction="harmonyPCA",ndims=50)

#run umap
srt=RunUMAP(srt,dims=1:30,reduction="harmonyPCA",reduction.name="harmonyUMAP",reduction.key="harmonyUMAP_")
#find nieghbor
srt=FindNeighbors(srt,dims=1:30,reduction="harmonyPCA",graph.name="h")
for(i in seq(0,2,0.25)){
  srt=FindClusters(srt,graph.name="h",resolution=i,verbose=FALSE)
}
library(clustree)
clustree(srt,prefix="h_res.")
```

```{r,fig.width=5,fig.height=5}
DimPlot(srt,reduction="harmonyUMAP",label=T,group.by="h_res.1.25") + NoAxes()
```


```{r,fig.width=20,fig.height=10}
VlnPlot(srt,features=c("nCount_RNA","nFeature_RNA","percent.mt","S.Score"),pt.size=0,ncol=2,group.by="h_res.1.25")
```

```{r,fig.width=15,fig.height=15}
DefaultAssay(srt)="DSB"
rownames(srt)[grep("CD1",rownames(srt))]
FeaturePlot(srt,reduction="harmonyUMAP",features=c(prot_to_use,"rna_GZMB"),min.cutoff="q2",max.cutoff="q98",ncol=4,order=T)
```






























end